---
title: "MIMIC-IV ICU Treatment Strategy RWE: Early Vasopressor vs No Early Vasopressor by 6 Hours"
author: "Xinyu Chen"
output: html_document
params:
  run_pipeline: false
  mimic_root: "/path/to/mimic-iv-3.1"
  outdir: "../outputs_reproduced"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(readr)
library(dplyr)
library(ggplot2)
library(scales)
project_root <- if (basename(getwd()) == "notebooks") normalizePath("..") else normalizePath(".")
find_output <- function(name) {
  candidates <- c(
    file.path(project_root, "outputs_reproduced", name),
    file.path(project_root, "outputs", name)
  )
  hit <- candidates[file.exists(candidates)][1]
  if (is.na(hit)) return(NA_character_)
  hit
}
```

# 1. Research question and groups compared

This project compares two ICU treatment-strategy groups among adult, first-ICU, sepsis-coded MIMIC-IV stays:

| Group | Operational definition |
|---|---|
| Treatment | First vasopressor initiation within 0--6 hours after ICU admission |
| Comparator | No vasopressor initiation within 0--6 hours after ICU admission |

The comparator is a **6-hour landmark comparator**, not a never-treated group. A patient who starts vasopressors after 6 hours belongs to the comparator under the primary strategy definition.

# 2. Target-trial-style design

| Element | Definition |
|---|---|
| Population | Adults, first ICU stay, sepsis-coded hospitalization, observed through the 6h landmark |
| Time zero | ICU admission time (`icustays.intime`) |
| Treatment | First vasopressor initiation within 0--6h |
| Comparator | No vasopressor initiation within 0--6h |
| Outcome | In-hospital mortality (`hospital_expire_flag`) |
| Baseline adjustment | Age, sex, admission type, first ICU unit, diagnosis count, first-6h vitals/labs |
| Main diagnostics | Table 1/SMD, propensity score overlap, love plot, crude vs adjusted results |

# 3. Why the final pipeline is consolidated

Earlier drafts separated extraction and analysis into different files. The later analysis-only file contained an important technical fix: after dummy encoding categorical covariates, the statsmodels weighted-logistic design matrix must be converted to numeric/float. Otherwise, bool/object columns can trigger model-fitting errors.

The final script `code/mimic_iv_early_vasopressor_pipeline.py` now combines both versions:

- extraction mode builds the cohort from raw MIMIC-IV CSV/CSV.GZ files;
- analysis-only mode reruns the downstream PS/IPTW/AIPW analysis from an existing cohort CSV;
- the dtype fix is integrated into the weighted logistic model section.

# 4. Cohort creation logic

The executable implementation is in `code/mimic_iv_early_vasopressor_pipeline.py`. The key cohort construction steps are reproduced below so the reader can see how the analysis dataset is created, not only how output CSV files are read.

```{sql, eval=FALSE}
-- 1) First ICU stay per patient
CREATE TEMP TABLE first_icu AS
SELECT * EXCLUDE(rn)
FROM (
  SELECT i.*, ROW_NUMBER() OVER (PARTITION BY subject_id ORDER BY intime) AS rn
  FROM icustays i
)
WHERE rn = 1;

-- 2) Diagnosis-code sepsis proxy
CREATE TEMP TABLE sepsis_hadm AS
SELECT DISTINCT hadm_id
FROM diagnoses_icd
WHERE
  (icd_version = 9 AND (icd_code LIKE '038%' OR icd_code IN ('99591','99592','78552')))
  OR
  (icd_version = 10 AND (icd_code LIKE 'A40%' OR icd_code LIKE 'A41%' OR icd_code LIKE 'R652%'));

-- 3) Adult first-ICU sepsis-coded base cohort
--    Time zero = ICU admission time.
--    The cohort excludes patients not observable through the 6h landmark.

-- 4) Vasopressor items are identified from ICU d_items labels:
--    norepinephrine, epinephrine, vasopressin, phenylephrine, dopamine.

-- 5) Treatment = first vasopressor start between ICU admission and ICU admission + 6h.
```

# 5. Run the pipeline from the notebook

Set `run_pipeline: true` in the YAML header or run the command below from the terminal. It is disabled by default because full local extraction can take time and requires credentialed MIMIC-IV files.

```{r run-pipeline, eval=params$run_pipeline}
cmd <- sprintf(
  'python "%s" --mimic_root "%s" --outdir "%s" --treatment_window_hours 6',
  file.path(project_root, "code", "mimic_iv_early_vasopressor_pipeline.py"),
  params$mimic_root,
  file.path(project_root, "outputs_reproduced")
)
system(cmd)
```

Analysis-only mode after extraction:

```{r analysis-only-command, eval=FALSE}
system(sprintf(
  'python "%s" --cohort_csv "%s" --outdir "%s" --analysis_only',
  file.path(project_root, "code", "mimic_iv_early_vasopressor_pipeline.py"),
  file.path(project_root, "outputs_reproduced", "mimic_icu_vasopressor_cohort.csv"),
  file.path(project_root, "outputs_reproduced")
))
```

# 6. Cohort flow and crude comparison

```{r cohort-flow}
cohort_path <- find_output("cohort_flow_summary.csv")
if (!is.na(cohort_path)) {
  cohort <- read_csv(cohort_path, show_col_types = FALSE)
  cohort
} else {
  cat("No cohort_flow_summary.csv found yet. Run the pipeline or inspect outputs/.")
}
```

```{r crude-rates}
crude_path <- find_output("crude_mortality_by_treatment_summary.csv")
if (!is.na(crude_path)) {
  crude <- read_csv(crude_path, show_col_types = FALSE)
  print(crude)
  crude %>%
    mutate(group_label = factor(group_label, levels = group_label)) %>%
    ggplot(aes(x = group_label, y = hospital_mortality_rate)) +
    geom_col() +
    geom_text(aes(label = percent(hospital_mortality_rate, accuracy = 0.1)), vjust = -0.3) +
    labs(x = "Treatment strategy group", y = "Crude hospital mortality", title = "Crude mortality by early vasopressor strategy") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 15, hjust = 1))
}
```

# 7. Propensity score and balance diagnostics

The pipeline estimates a propensity score for early vasopressor initiation using pre-exposure or first-6h baseline variables. It then computes stabilized IPTW and trims extreme weights at the 1st and 99th percentiles.

```{r balance-summary}
bal_path <- find_output("balance_summary.csv")
if (!is.na(bal_path)) {
  read_csv(bal_path, show_col_types = FALSE)
} else {
  cat("No balance_summary.csv found yet.")
}
```

If reproduced locally, inspect:

```text
outputs_reproduced/propensity_overlap.png
outputs_reproduced/love_plot.png
outputs_reproduced/table1_smd_before_after.csv
```

# 8. Adjusted estimates and interpretation

```{r aipw-results}
aipw_path <- find_output("aipw_results_summary.csv")
if (!is.na(aipw_path)) {
  read_csv(aipw_path, show_col_types = FALSE)
} else {
  cat("No AIPW summary found yet.")
}
```

The crude mortality contrast should be interpreted cautiously. Early vasopressor initiation is also a marker of shock severity and clinical urgency. The adjusted analysis is designed to show how a target-trial-style workflow can reduce observable imbalance; it does not prove a randomized causal effect.
